GravExplain: Continuous gravitational wave searches in a table-top experiment

### 2020-02-13 paper/paper_main.tex
setting up merge conflict

% cover page
\section*{Project contributions}
% contributions clarify what’s to be assessed
James W. Gardner:
\begin{itemize}
\item all paper sections other than the introduction, i.e.\ from Section~\ref{sec:ifo} onwards, and all figures in the paper
\item addition of the injection speaker, design of photodiode circuit from standard examples, assembly of circuit on breadboard, and set-up of Raspberry Pi to measure output
\item all of the source code at \url{https://github.com/daccordeon/gravexplain}, that reads the data from the Raspberry Pi, analyses it, and creates the figures used in the paper
\item talks presented were based on paper and figures therein
\end{itemize}

@@

\subsection*{Statement of achievements}
% one paragraph recount of what’s been done for two months, like the abstract

\jam{write this!}

@@

% AJP requires an abstract for all regular article submissions.
\begin{abstract}
Continuous gravitational wave searches use sophisticated statistical techniques to search for signals in noisy data. We demonstrate some of these techniques with a table-top Michelson interferometer, namely the Viterbi algorithm for recovering a tone with wandering frequency.
We also use the same experiment to demonstrate an optical microphone capable of playing back simple audio recordings of injected music. However, played-back speech is currently unintelligible so we investigate preliminary methods of speech enhancement.

\jam{Andrew says that this should be a bit more detailed!}

@@

\section{Complex audio}
\label{sec:optical_microphone}
% Optical microphone
% aim of optical microphone: play back injected sounds

The natural succession to simple tones is to study complex audio signals, such as music and speech.
The Viterbi algorithm and tracking more broadly is not applicable to these signals. Compare a single tone that slowly, continuously changes frequency to music and speech which consist of a broad spectrum that is constantly changing.
Instead, the interest here is in being able to record and play-back these complex audio signals, rather than track them. Using the interferometer to do so makes it, in ensemble with the measuring device, an ‘optical microphone’, using light to capture sound.

@@

For an optical microphone to be sensitive to speech and music it needs a frequency response far higher than the 30Hz of a standard webcam. Both speech and music need at least kHz range, with speech intelligibility requiring up to 3kHz and music preferably up to and beyond 8kHz. This can be addressed by using a high-speed webcam, or more simply, by using a photodiode as the measuring device.

@@

The voltage signal from the photo-detector is captured by a 10-bit analog-to-digital converter (ADC) connected to the special purpose input (SPI) pins of a Raspberry Pi, in the standard configuration. This digitally samples the signal at roughly 16kHz. As such, any frequency component of the analog signal above the Nyquist frequency of 8kHz would be aliased down into the detected range.
To prevent this, an anti-aliasing Sallen-Key filter tuned to 16kHz is included before the ADC. This component attenuates any frequencies above 8kHz, before they get digitally sampled.
It’s not immediately clear what’s limiting the sample rate to 16kHz, but it is likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figures/setup_pic2.pdf}
	\caption{Picture taken of the optical microphone set-up, showing the Michelson interferometer, the photodiode (in an insert), the circuit, and the Raspberry Pi.}
	\label{fig:setup_pic2}
\end{figure}

@@

\subsection{Filters applied}

@@

Note that we also replicated the Viterbi algorithm results from before with the photodiode, but only with the mains hum removed and high-passing above 100Hz, else the algorithm never selected the signal. Except for being over audible frequencies, there isn’t much difference in the tracking once these filters are applied.

@@

\subsection{logMMSE estimator}
% logMMSE

@@

\subsection{Transfer function}

@@

For the photodiode circuit, we used a Raspberry Pi and an MCP3008 ADC to monitor the analog signal. This could also be done with the analog pins on an Arduino. Be careful of the gain through the LM358 op-amps in the circuit though, else the ADC in either case could threshold the signal. Also be ready to change the values in the Sallen-Key filter to match your own sampling rate. The photodiode was an OSRAM BPW21, we mounted it through a discarded piece of speaker mesh (similar to black cloth over a frame) and placed another piece over the face to not over expose and threshold the signal. For the rest of the circuit, we used standard components assembled as in \ref{app:circuit_diagram}.

\jam{Add this in, lots of advice and detail!}

\section{The Viterbi Algorithm}
\label{app:viterbi}

@@

\section{Common wandering frequency mistake}
\label{app:phase_gotcha}

@@

Figure~\ref{fig:circuit_diagram} shows the photodiode circuit diagram, it is also available at \url{https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce}

\begin{figure}%[H]
	\includegraphics[width=\textwidth]{figures/circuit_diagram_2.pdf}
	\caption{Circuit diagram for photodiode reading. Photodiode in reverse-bias over an op-amp, analog signal then passed through Sallen-Key anti-aliasing filter tuned to 16kHz, then into an analog-to-digital converter (ADC), that’s finally read by the special purpose input (SPI) pins of a Raspberry Pi}
	\label{fig:circuit_diagram}
\end{figure}

>>>>

% cover page
\section*{Project contributions}
% contributions clarify what’s to be assessed
James W. Gardner:
\begin{itemize}
\item all paper sections other than the introduction, i.e.\ from Section~\ref{sec:ifo} onwards, and all figures in the paper
\item addition of the injection speaker, design of photodiode circuit from standard examples, assembly of circuit on breadboard, and set-up of Raspberry Pi to measure output
\item all of the source code at \url{https://github.com/daccordeon/gravexplain}, that reads the data from the Raspberry Pi, analyses it, and creates the figures used in the paper
\item talks presented were based on the paper and figures therein
\end{itemize}

@@

\subsection*{Statement of achievements}
% one paragraph recount of what’s been done for two months, like the abstract
The project spanned three weeks in December, four weeks in January, and two weeks in February. The first two weeks were spent writing the initial scripts and achieving the Viterbi results with the webcam set-up. The third week was spent setting up the photodiode circuit and the Raspberry Pi. All of January was spent progressing through the optical microphone experiment, and was largely driven by suggestions made at the signal processing meetings with the electrical engineers. By the end of the third week of January we had a working optical microphone but low speech intelligibility. The fourth week of January was spent on attempting methods of speech enhancement. Finally, the two weeks in February were spent writing and drafting the paper/report, and giving presentations.

@@

% AJP requires an abstract for all regular article submissions.
\begin{abstract}
Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. We demonstrate some of these techniques by observing sound with a table-top Michelson interferometer. In particular, we successfully use the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
We also demonstrate how the inclusion of a photodiode and Raspberry Pi allows the interferometer to act as an optical microphone. We are able to play-back simple audio recordings from the photodiode data. However, played-back speech is unintelligible and so we investigate techniques of speech enhancement.

@@

\section{Playing back complex audio}
\label{sec:optical_microphone}
% Optical microphone
% aim of optical microphone: play back injected sounds

The natural succession to observing simple tones is to study complex audio signals, such as music and speech.
The Viterbi algorithm and tracking more broadly is not applicable to these signals. Compare a single tone that slowly, continuously changes frequency to music and speech which consist of a broad spectrum that is constantly changing.
Instead, the interest here is in being able to record and play-back these complex audio signals, rather than track them. Using the interferometer to do so makes it, in ensemble with the measuring device, an ‘optical microphone’, using light to capture sound.

@@

For an optical microphone to be sensitive to speech and music it needs a frequency response (a sampling rate) far higher than the 30Hz of a standard webcam. Both speech and music need at least kHz range, with speech intelligibility requiring up to 3kHz and music preferably up to and beyond 8kHz\cite{speech_intelligibility}. This can be addressed by using a high-speed webcam, or more simply, by using a photodiode as the measuring device.

@@

The voltage signal from the photo-detector is captured by a 10-bit analog-to-digital converter (ADC) connected to the special purpose input (SPI) pins of a Raspberry Pi, in the standard configuration. This digitally samples the signal at roughly 16kHz. As such, any frequency component of the analog signal above the Nyquist frequency of 8kHz would be aliased down into the detected range.
To prevent this, an anti-aliasing Sallen-Key filter\cite{sallen_key_filter} tuned to 16kHz is included before the ADC. This component attenuates any frequencies above 8kHz, before they get digitally sampled.
It’s not immediately clear what’s limiting the sample rate to 16kHz, but it is likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figures/setup_pic2.pdf}
	\caption{Picture taken of the optical microphone set-up, showing the Michelson interferometer, the photodiode (in an insert), the circuit, and the Raspberry Pi. For a closer photo of the circuit, see Figure~\ref{fig:circuit_pic3}.}
	\label{fig:setup_pic2}
\end{figure}

@@

\subsection{Applying filters to the data}

@@

% Note that we also replicated the Viterbi algorithm results from before with the photodiode, but only with the mains hum removed and high-passing above 100Hz, else the algorithm never selected the signal. Except for being over audible frequencies, there isn’t much difference in the tracking once these filters are applied.

@@

\subsection{The logMMSE estimator}
% logMMSE

@@

\subsection{Finding the transfer function}

@@

For the photodiode circuit, we used a Raspberry Pi and an MCP3008 ADC to monitor the analog signal. This could also be done with the analog pins on an Arduino. Be careful of the gain through the LM358 op-amps in the circuit though, else the ADC in either case could threshold the signal. Also be ready to change the values in the Sallen-Key filter to match your own sampling rate. The photodiode was an OSRAM BPW21, we mounted it through a discarded piece of speaker mesh (similar to black cloth over a frame) and placed another piece over the face to not over expose and threshold the signal. For the rest of the circuit, we used standard components assembled on a breadboard as detailed in Appendix~\ref{app:circuit_diagram}.

\section{The Viterbi algorithm}
\label{app:viterbi}

@@

\section{A common mistake with wandering frequency signals}
\label{app:phase_gotcha}

@@

Figure~\ref{fig:circuit_diagram} shows the photodiode circuit diagram, it is also available at \url{https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce}. A photo of the breadboard and Raspberry Pi pins is shown in Figure~\ref{fig:circuit_pic3}.

\begin{figure}%[H]
	\includegraphics[width=\textwidth]{figures/circuit_diagram_2.pdf}
	\caption{Circuit diagram for photodiode reading. Photodiode in reverse-bias over an op-amp, analog signal then passed through Sallen-Key anti-aliasing filter tuned to 16kHz, then into an analog-to-digital converter (ADC). The digitised signal is then read by the special purpose input (SPI) pins of a Raspberry Pi in standard configuration.}
	\label{fig:circuit_diagram}
\end{figure}

\begin{figure}%[H]
	\includegraphics[width=0.8\textwidth]{figures/circuit_pic3.pdf}
	\caption{Photo of circuit for photodiode reading assembled on breadboard.}
	\label{fig:circuit_pic3}
\end{figure}
